D3064R0: OOTA Execution is Provably Vacuous

### 2023-12-11 ootavacuous.tex
December 10 self-review
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

OOTA occurs when a group of threads load from each others' stores,
but where each thread's store depends on the value returned by that
thread's load.
A given out-of-thin-air value passes around the resulting cycle.
Because there is a cycle, on a real system, at least one of the links
must be counter-temporal, that is, either a load returns a value before
that value is stored, or a store's address or value is computed before
a prior load returns a value used in that computation.

@@

Line~4 of the corresponding output of the \co{herd7} tool shows the
counterintuitive outcome where both processes load the value 42.
As we will see, this outcome is not possible in real-world implementations
due to physical constraints of which \co{herd7} is (by design) unaware.

@@

Some executable C++ memory models correctly flag at least some executions
involving OOTA cycles~\cite{JadeAlglave2014HerdingCats}.\footnote{
	Others cleverly avoid this issue by forbidding atomic
	stores of non-constant values~\cite{MarkBatty2011cppmem}.}
% @@@ Better herd7 C11 citation?
However, because these models are atemporal, they cannot reject
OOTA executions other than by flagging the OOTA value as arbitrary
(which some in fact do in at least some cases).

@@

Goldblatt looked at interactions between OOTA cycles and
UB~\cite{DavidGoldblatt2019NoElegantOOTAfix}.
Appendix~\ref{app:Aside on Undefined Behavior}
analyzes the examples and proposes a way to separate UB and OOTA-cycle
concerns.
The remainder of this document will focus on examples lacking UB.

@@

Others, current authors included, have define an OOTA cycle as a
cycle in sdep $\cup$ rf, where sdep is the set of semantic
dependencies within each thread and rf is the set of store-to-load
links, whether intra-thread or between threads.
This is a fine definition, and is consistent with the words in the C++
standard, except that it simply shunts all the complexity into the
term ``semantic dependency''.

@@

Similarly:

@@

These examples demonstrate a key point: Although in some cases sdep can
be a strict function of the source code, \emph{in general, sdep must be
defined on a per-execution basis}.

@@

Fortunately, there are now tools that can help.
For example, the SAT-solver-based \co{cbmc} tool
has been used to mechanically verify signicant portions of Linux-kernel
RCU from the C-language source
code~\cite{LihaoLiang2016VerifyTreeRCU,LanceRoy2017CBMC-SRCU}.
Another tool, Nidhugg, which is based on partial-order
reduction, has been used to carry out a similar
verification~\cite{MichalisKokologiannakis2017NidhuggRCU,SMC-TreeRCU,MichalisKokologiannakis2019RCUstatelessModelCheck}.
Both tools can easily check whether or not a given dependency is semantic.
An example use of \co{cbmc} is illustrated in
Appendix~\ref{sec:Evaluating sdep Using cbmc}.

@@

All of these tools can incur high overhead, especially manual code
inspection.
However, as we will see, the actual C++ implementations themselves need
not use these tools.
Their current analysis is guaranteed to be sufficient, courtesy of the
real-world constraints discussed in the next section.
Furthermore, as we will also see, in some special but commonly occurring
circumstances, there are much simpler methods that incur negligible overhead.

@@

This paper will therefore assume that the speed of light is finite, and
thus the maximum speed at which information can propagate is also finite.

@@

If atoms were of zero size, a zero-sized computer might be constructed.
In such a computer, even give finite speed of light, information could
propagate among computing elements in zero time.
This situation would also permit a load to return the value stored by
some other thread at the exact instant that the store executed, which
would again in turn remove propagation delay as a factor preventing
causal loops such as OOTA cycles from forming.

@@

These three laws of physics mean that if one thread loads the value
stored by some other thread, that load must have occurred later
in global time than did the store.
In other words, the reads-from relation is temporal in nature.

@@

Applying this to Listing~\ref{lst:OOTA Cycle}, because values from
speculative stores cannot be visible to non-speculative loads, lines~10
and~12 cannot be committed until after the load in line~8 completes
and the conditional on line~9 is evaluated non-speculatively.\footnote{
	Alternatively, one could imagine that both the speculation
	and the squashing of failed speculation might span multiple
	threads.}
Similarly, lines~18 and~20 cannot be committed until after lines~16
and~17 complete non-speculatively.

@@

However, if all of the inter-thread relaxed accesses are volatile,
as they are in the C language, then all of the inter-thread traffic is
observed behavior.
In this case, even omniscient oracular implementations are required
to confine their analyses within a single thread.

@@

\begin{table}
\centering
\begin{tabular}{c|c|l}
Actual Code	& Implementation	& Result \\
\hline
sdep		& sdep			& Temporal \\
\cline{2-3}
		& $\neg$sdep		& Fix C++ bug! \\
\hline
$\neg$sdep	& sdep			& Temporal \\
\cline{2-3}
		& $\neg$sdep		& Atemporal \\
\end{tabular}
\caption{Semantic-Dependency Classification}
\label{tab:Semantic-Dependency Classification}
\end{table}

@@

\begin{table}
\centering
\begin{tabular}{c|c|l}
Actual Code	& Tooling		& Result \\
\hline
sdep		& sdep			& Temporal \\
\cline{2-3}
		& $\neg$sdep		& False Positive (Warning?) \\
\hline
$\neg$sdep	& sdep			& False Negative (Warning?) \\
\cline{2-3}
		& $\neg$sdep		& Atemporal \\
\end{tabular}
\caption{Semantic-Dependency Classification For Tools}
\label{tab:Semantic-Dependency Classification For Tools}
\end{table}

@@

We are grateful to David Goldblatt, Jade Alglave, and Alan Stern for their
careful review of an early draft of this paper and to John Wickerson for
asking Paul for a rant and taking the proffered rant seriously.
We also owe David Goldblatt a debt of gratitude for his having asked an
insightful question at the right time, his insights on combinations of
OOTA and UB, and for his ``Deathstation 9000'' demonic CPU.
Alan Stern located many unclear statements and gaps in reasoning.
Martin Uecker contributed valuable insights on backwards-propagating UB.
Nonetheless, all errors and omissions in this paper are the sole property
of the authors.

@@

% @@@ TODO:
% Measure store-buffer delays on an x86 system.

>>>>

OOTA occurs when a group of threads load from each others' stores,
but where each thread's store depends on the value returned by that
thread's load.
A given out-of-thin-air value passes around the resulting cycle.
Because there is a cycle, on a real system, at least one of the links
must be atemporal, that is, either a load returns a value before
that value is stored, or a store's address or value is computed before
a prior load returns a value used in that computation.

@@

Listing~\ref{lst:OOTA Cycle, herd7 Output} shows the corresponding output
of the \co{herd7} tool.
Lines~3 and~4 show the possible states, with line~4 showing the
counterintuitive outcome where both processes load the value 42.
As we will see, this outcome is not possible in real-world implementations
due to physical constraints of which \co{herd7} is (by design) unaware.
Line~9 contains \co{Sometimes} (as opposed to \co{Never} or \co{Always}),
indicating that some executions satisfy the \co{exists} clause and
others do not.

@@

Some executable C++ memory models correctly flag at least some executions
involving OOTA cycles~\cite{JadeAlglave2014HerdingCats}.\footnote{
	Others cleverly avoid this issue by forbidding atomic
	stores of non-constant values~\cite{MarkBatty2011cppmem}.}
% @@@ Better herd7 C11 citation?
However, because these models are atemporal, they cannot reject
OOTA executions other than by flagging the OOTA value as arbitrary,
which some in fact do in at least some cases.

@@

Goldblatt looked at interactions between OOTA cycles and
UB~\cite{DavidGoldblatt2019NoElegantOOTAfix}.
Appendix~\ref{app:Aside on Undefined Behavior}
analyzes the examples and notes ways to separate UB and OOTA-cycle
concerns.
The remainder of this document will focus on examples lacking UB.

@@

Others, current authors included, have defined an OOTA cycle as a
cycle in sdep $\cup$ rf, where sdep is the set of semantic
dependencies within each thread and rf is the set of store-to-load
links, whether intra-thread or between threads.
This is a fine definition, and is consistent with the words in the C++
standard, except that it simply shunts all the complexity onto the
term ``semantic dependency''.

The following sections discuss properties of and constraints on
useful definitions of ``semantic dependency''.

@@

Consider also:

@@

These examples demonstrate a key point: Although in some cases sdep can
be a strict function of the source code, \emph{in general, sdep must be
defined on a per-execution basis}.\footnote{
	Thanks to Alan Stern for the examples in this section and
	in the preceding one.}

\subsubsection{Semantic Dependencies Can Be Nondeterministic}
\label{sec:Semantic Dependencies Can Be Nondeterministic}

Consider this example, with \co{i} initially zero:

\begin{quote}
	\co{x = ++i / ++i;}
\end{quote}

The standard does not specify the order of evaluation, so the resulting
value might be zero ($\frac{1}{2}$ truncated) or two ($\frac{2}{1}$).
This indeterminacy is why many coding guidelines prohibit side effects
in expressions, but the above code nevertheless conforms to the standard.

@@

Fortunately, there are now tools that can help.
For example, the SAT-solver-based \co{cbmc} tool
has been used to mechanically verify signicant portions of Linux-kernel
RCU from the C-language source
code~\cite{LihaoLiang2016VerifyTreeRCU,LanceRoy2017CBMC-SRCU}.
Another tool, Nidhugg, which is based on partial-order
reduction, has been used to carry out a similar
verification~\cite{MichalisKokologiannakis2017NidhuggRCU,SMC-TreeRCU,MichalisKokologiannakis2019RCUstatelessModelCheck}.
Both tools can easily check whether or not a given dependency is semantic
for at least some approximate definitions of ``semantic dependency''.
An example use of \co{cbmc} is illustrated in
Appendix~\ref{sec:Evaluating sdep Using cbmc}.

@@

All of these tools can incur high overhead, especially manual code
inspection.
However, as we will see, the actual C++ implementations themselves need
not use these tools.
Their current analysis is guaranteed to be sufficient, courtesy of the
real-world constraints discussed in the next section, at least assuming
that their analysis is in fact correct.
Furthermore, as we will also see, in some special but commonly occurring
circumstances, there are much simpler methods that incur negligible overhead.

@@

This paper will therefore assume that the speed of light is finite, and
that the maximum speed at which information can propagate is also finite.

@@

If atoms were of zero size, a zero-sized computer might be constructed.
In such a computer, even given finite speed of light, information could
propagate among computing elements in zero time.
This situation would also permit a load to return the value stored by
some other thread at the exact instant that the store executed, which
would again in turn remove propagation delay as a factor preventing
causal loops such as OOTA cycles from forming.

@@

These three laws of physics mean that if one thread loads the value
stored by some other thread, that load must have occurred earlier
in global time than did the store.
In other words, the reads-from relation is temporal in nature.

@@

Applying this to Listing~\ref{lst:OOTA Cycle}, because values from
speculative stores cannot be visible to non-speculative loads, lines~10
and~12 cannot be committed until after the load in line~8 completes
and the conditional on line~9 is evaluated non-speculatively.\footnote{
	Alternatively, one could imagine that both the speculation
	and the squashing of failed speculation might span multiple
	threads.
	This might make sense for multiple hardware threads sharing
	a core.}
Similarly, lines~18 and~20 cannot be committed until after lines~16
and~17 complete non-speculatively.

@@

However, if all of the inter-thread relaxed accesses are volatile,
as they are in the C language, then all of the inter-thread traffic is
observed behavior.
In this case, even omniscient oracular implementations are required
to confine their analyses to within a single thread.

@@

\begin{table}
\centering
\begin{tabular}{c|c|l}
Actual Execution	& Implementation	& Result \\
\hline
sdep			& sdep			& Temporal \\
\cline{2-3}
			& $\neg$sdep		& Fix C++ bug! \\
\hline
$\neg$sde		& sdep			& Temporal \\
\cline{2-3}
			& $\neg$sdep		& Atemporal \\
\end{tabular}
\caption{Semantic-Dependency Classification}
\label{tab:Semantic-Dependency Classification}
\end{table}

@@

\begin{table}
\centering
\begin{tabular}{c|c|l}
Actual Execution	& Tooling		& Result \\
\hline
sdep			& sdep			& Temporal \\
\cline{2-3}
			& $\neg$sdep		& False Positive (Warning?) \\
\hline
$\neg$sdep		& sdep			& False Negative (Warning?) \\
\cline{2-3}
			& $\neg$sdep		& Atemporal \\
\end{tabular}
\caption{Semantic-Dependency Classification For Tools}
\label{tab:Semantic-Dependency Classification For Tools}
\end{table}

@@

We are grateful to David Goldblatt, Jade Alglave, Alan Stern, and Peter
Sewell for their careful review of an early draft of this paper and
to John Wickerson for asking Paul for a rant and taking the proffered
rant seriously.
We also owe David Goldblatt a debt of gratitude for his having asked an
insightful question at the right time, his insights on combinations of
OOTA and UB, and for his ``Deathstation 9000'' demonic CPU.
Alan Stern located many unclear statements and gaps in reasoning.
Martin Uecker contributed valuable insights on backwards-propagating UB.

Nonetheless, all errors and omissions in this paper are the sole property
of the authors, and the appearance of a name in this paragraph does
not in any way constitute agreement with anything in this paper.

@@

% @@@ TODO:
% Measure store-buffer delays on an x86 system.
% Real-world compilers freeze nondeterministic choices.
